---
title: "Lab 1 Introduction to R Markdown"
author: "B. Sosnovski"
date: "09/25/2021"
output: 
  html_document: 
    theme: cerulean
---
```{r setup, include=FALSE, warning = FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction to R Markdown

This lab will show you how to write an R Markdown document.

## General Goals

In this lab you will learn:

1. What R Markdown is
2. How to use of most commonly R Markdown syntax
3. How to create an R Markdown file
4. How to add code in your R Markdown file
5. How to format your text and equations in R Markdown

## Markdown

Markdown is a way of adding to a plain text document formatting such as headers, lists, emphasis, quotes, links to URLs and images, and more.

Markdown allows you to write using an easy-to-read, easy-to-write plain text format, then convert it to HTML.

## R Markdown

R Markdown allows you to write a  HTML, PDF, and/or MS Word documents, a presentation or even a book by just using Markdown syntax and embedded programming code.

R Markdown files have extension **.Rmd**

The "**R**" in R Markdown stands for the [R statistical computing](https://www.R-project.org/) language.

R Markdown files contain three important types of content:

1. An (optional) YAML header surrounded by ---s.
2. Chunks of R code surrounded by ```.
3. Text mixed with simple text formatting like # heading and _italics_.

We will cover these contents in this lab.


## Load Packages

Packages are collections of R functions, data, and some compiled code in a well-defined format. The directory where packages are stored is called the library. R comes with a standard set of packages.  Others packages are available for download and  must be installed. 

In Cocalc, a large number of packages are already installed, so there is no need to install them.

Once a package is installed, they have to be loaded into the session to be used.

To load a package in R, we invoke the command ``library(package)`` where in ''package'' will be replaced with the name of the package you want to load.

Here is an example on how to invoke a package in R Markdown.

```{r}
library("knitr")
```

The ``knitr`` package extends the basic markdown syntax to include chunks of executable R code.

Cocalc already have this package loaded for you. But if you are running your code, say in R Studio, you would need to install and include the ``knitr`` in the library of your document or script.


### Example

Now load the package called ``dplyr``.

`Enter your code below:`


You will learn how to install packages for R in Cocalc in lab 2 (Stay tuned!!!).


## Formatting text

I hope you already noticed some of the formatted text above in this rmd file.

Here some basics formatting of text in Markdown.

### Comments

Note that you can add comments in Markdown documents as follows: ``<!-- your comment -->``. The comments will not be displayed in any output format.

<span style='color:#d2691e'>Compare this line here in the rmd file on the left and the output format in HTML on the right (Note that the comment below is missing!!.</span>

<!--  This will not be displayed in the output format -->

### Plain Code Blocks

Plain code blocks are displayed in a fixed-width font but not evaluated by R.

```
2*3
This text is displayed verbatim
```
You will see this throughout this rmd file to display examples of codes that are not evaluated.


### Headers
   
   You can include section headers by using pound signs.
   

```
# First-level header

## Second-level header

### Third-level header
```

### Emphasis


```
*italic*   **bold**

_italic_   __bold__

```
Note that the word *Statistics* is in italic here and the work **markdown** is in bold.

*Bianca*
**Sosnovski**
  
### Lists

##### Unordered list

* Item 1
* Item 2
    + Item 2a
    + Item 2b

The code for the list above is:

```
* Item 1
* Item 2
    + Item 2a
    + Item 2b
    
```
* Grade test 1
* Grade HW
    + HW 1
    + HW 2


##### Ordered list

1. Item 1
2. Item 2
3. Item 3
    + Item 3a
    + Item 3b


The code for the list above is:

```
1. Item 1
2. Item 2
3. Item 3
    + Item 3a
    + Item 3b
```
### Hyperlinks

To add hyperlinks we surround the links with brackets, and then provide the link address in parentheses, like this [QCC](https://www.qcc.cuny.edu).

```
[QCC](https://www.qcc.cuny.edu)
```

## Embedding R code


You can embed an R code chunk like this:

```{r}
summary(cars)
#cars is a built-in data set in R
```

Note that the line above is a comment in the R code, and that you can add comments in the R code chunk with a pound sign. Another example:

```{r}
2*3
# This is purely a comment that will be ignored by the R evaluator.
```

To add an inline code use a pair of backticks and the letter r. For example, `r 2*3`. We can also mark text as inline code by using just the pair of backticks. For example, `code`.

You can also embed plots in your rmd files. For example:

```{r, echo=FALSE}
plot(cars)
```

You can do a lot of things in a code chunk: produce text output, tables, or graphics. You have fine control over all these output via chunk options, which can be provided inside the curly braces (between `{r and }`).

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### Example
Create a code chunk that will produce the summary of the dataset `Seatbelts`.

`Enter your code below:`


### Example
Create a code chunk that will produce a plot of the dataset `Seatbelts`.

`Enter your code below:`



## Writing mathematical equations and formulas

Of course, you can include math formulas by using dollar signs around the equation in Latex syntax (Using $ symbols around the equation).

The following is the code for writing the equation of the area of a circle in two formats.

### In line

```
$ A= \pi r^2 $
``` 

### In equation display

```
$$ A= \pi r^2 $$ 
```

### Example

Write the equation of the circumference of a circle. 

`Enter your code below:`


## Lab Exercises and Writing Assignment

The instructions for the lab and WA are in the Lab_1.rmd and WA_1.rmd files, which will be used to submit your work and answers.

## References

Parts of the material used in this document are adaptations or excerpt from the following:

* Grolemund, Garrett. "Introduction to R Markdown."  July 16, 2014. https://rmarkdown.rstudio.com/articles_intro.html

* "Markdown Basics." https://rmarkdown.rstudio.com/authoring_basics.html

